---
title: "Cost of Retracted Articles to the NIH: a Living Analysis"
author:
  - "Alejandro Sandoval-Lentisco^1^"
  - "John P.A. Ioannidis^1^^2^^3^^4^"
format:
  html:
    toc: false
    number-sections: false
    css: |
      h2 { text-transform: uppercase; letter-spacing: 0.05em; font-size: 1.05em; }
      .analysis-tag { letter-spacing: 0.15em; font-weight: 700; color: #666; }
  pdf:
    documentclass: article
    geometry: margin=2cm
  docx: default
execute:
  echo: false
  warning: false
  message: false
editor_options:
  chunk_output_type: inline
bibliography: references.bib
csl: science.csl
---

  ^1^ Meta-Research Innovation Center at Stanford (METRICS), Stanford University, Stanford, CA, USA
  ^2^ Stanford Prevention Research Center, Department of Medicine, Stanford University, Stanford, CA, USA
  ^3^ Department of Epidemiology and Population Health, Stanford University, Stanford, CA, USA
  ^4^ Department of Biomedical Data Science, Stanford University, Stanford, CA, USA


**Correspondence:** John P.A. Ioannidis, MD, DSc, 3180 Porter Drive, Stanford Research Park, Palo Alto, CA 94304, USA. E-mail: jioannid\@stanford.edu

```{r setup, include=FALSE}
# Setup, libraries, and data loading
library(tidyverse)
library(scales)
library(lubridate)
library(ggplot2)
library(patchwork)
library(gridExtra)

# Shared colour scheme and ggplot theme used by every figure in this document (and
# matched in supplementary_materials.qmd and app.R so all outputs look identical):
# orange = NIH-funded, green = not NIH-funded.
corporate_colors <- c("NIH Funded" = "#d16b2a",
                      "Not NIH Funded" = "#00716f")

dashboard_theme <- theme_classic(base_size = 18) +
  theme(
    legend.position  = "top",
    legend.title     = element_blank(),
    legend.text      = element_text(size = 20),
    axis.title       = element_text(size = 20, face = "bold"),
    axis.text        = element_text(size = 18),
    plot.background  = element_rect(fill = "transparent", color = NA),
    panel.background = element_rect(fill = "transparent", color = NA),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.5)
  )

# Input files
papers_file  <- "data/processed/FINAL_Retractions_Costs_and_Pubs.csv"
authors_file <- "data/processed/Author_Funding_Matches.csv"
cpi_file     <- "data/processed/annual_cpi.csv"
collision_flags_file <- "data/processed/author_collision_flags.csv"  # built by build_collision_flags.R
grant_keep_file      <- "data/processed/author_grant_keep.csv"       # built by build_collision_flags.R

# Load CPI
annual_cpi <- read_csv(cpi_file, show_col_types = FALSE)
if (!"Inflation_Multiplier" %in% colnames(annual_cpi)) {
  current_cpi <- max(annual_cpi$CPI, na.rm = TRUE)
  annual_cpi$Inflation_Multiplier <- current_cpi / annual_cpi$CPI
}
cpi_year <- max(annual_cpi$Fiscal_Year, na.rm = TRUE)

# Load Papers. One row per retracted article. From each row we derive:
#   Publication_Year / Retraction_Year -> the 4-digit year parsed from the date strings;
#   retraction_lag       -> years between publication and retraction;
#   Funding_Status       -> "NIH Funded" if the paper has any NIH grant cost, else not;
#   attributed_cost_raw  -> the paper's share of its grants' cost = total NIH funding
#                           divided by the number of publications produced under those
#                           grants (a per-paper attribution, in nominal dollars).
# The CPI join then adds attributed_cost_adjusted (the raw cost in cpi_year dollars).
df_papers <- read_csv(papers_file, show_col_types = FALSE) %>%
  mutate(
    Publication_Year = as.numeric(str_extract(OriginalPaperDate, "(?<=/)\\d{4}")),
    Retraction_Year  = as.numeric(str_extract(RetractionDate,    "(?<=/)\\d{4}")),
    retraction_lag   = Retraction_Year - Publication_Year,
    Funding_Status   = ifelse(!is.na(Total_NIH_Funding_Cost) & Total_NIH_Funding_Cost > 0,
                              "NIH Funded", "Not NIH Funded"),
    attributed_cost_raw = ifelse(
      !is.na(Total_Grant_Pubs) & Total_Grant_Pubs > 0,
      Total_NIH_Funding_Cost / Total_Grant_Pubs,
      NA
    )
  ) %>%
  left_join(annual_cpi, by = c("Publication_Year" = "Fiscal_Year")) %>%
  mutate(attributed_cost_adjusted = attributed_cost_raw * Inflation_Multiplier)

# Load Authors
df_authors_raw <- read_csv(authors_file, show_col_types = FALSE)

author_retractions <- df_authors_raw %>%
  group_by(Original_Author_Name) %>%
  summarise(Num_Retractions = n_distinct(Record_ID), .groups = "drop")

# Grant-level name-collision filter (built by build_collision_flags.R). Common author names can
# match the grant records of several different NIH investigators; for an author whose name
# collides with more than one investigator we keep only the grants belonging to the real person's
# PI_ID (identified via the institution on their retracted paper; PI_ID is stable across
# institution moves) and drop grants held by same-named others. Non-collision authors keep every
# grant. See the Supplementary methods for the full procedure and a sensitivity comparison.
grant_keep <- read_csv(grant_keep_file, show_col_types = FALSE)

df_authors <- df_authors_raw %>%
  group_by(Original_Author_Name) %>%
  mutate(First_Retraction_Year = min(Retraction_Year, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(diff_retracted = Fiscal_Year - First_Retraction_Year) %>%
  left_join(grant_keep, by = c("Original_Author_Name", "Grant_ID")) %>%
  filter(coalesce(keep_consequences, TRUE)) %>%
  distinct(Original_Author_Name, First_Retraction_Year, Grant_ID, Fiscal_Year, Funding_Amount, diff_retracted) %>%
  left_join(annual_cpi, by = "Fiscal_Year") %>%
  mutate(Funding_Adjusted = Funding_Amount * Inflation_Multiplier) %>%
  left_join(author_retractions, by = "Original_Author_Name")

# Restrict the author-consequences cohort to authors whose first retraction is early
# enough that a full +3-year post-retraction window is observable.
latest_fiscal_year     <- max(df_authors$Fiscal_Year, na.rm = TRUE)
retraction_cutoff_year <- latest_fiscal_year - 3
df_authors_cohort      <- df_authors %>% filter(First_Retraction_Year <= retraction_cutoff_year)

# Descriptive counts for the name-collision correction (the filter itself is applied above, at
# the grant level, via the grant_keep join). n_collision_authors: author names mapping to >1
# NIH investigator. n_excluded_collision: collision authors in the consequences window who have
# no institution-matching grant and therefore lose all attributed funding and leave the cohort.
collision_flags      <- read_csv(collision_flags_file, show_col_types = FALSE)
n_collision_authors  <- sum(collision_flags$n_pi_id > 1)
n_cohort_unfiltered  <- df_authors_raw %>% group_by(Original_Author_Name) %>%
  summarise(FRY = min(Retraction_Year, na.rm = TRUE), .groups = "drop") %>%
  filter(FRY <= retraction_cutoff_year) %>% nrow()
n_excluded_collision <- n_cohort_unfiltered - n_distinct(df_authors_cohort$Original_Author_Name)

# Helpers
dollar_short <- scales::label_dollar(accuracy = 0.01, scale_cut = scales::cut_short_scale())
# Dollar string with the leading "$" escaped as "\$". A bare "$amount" placed just before inline
# math such as "($V$ = ...)" lets pandoc pair the two stray "$" into a spurious math span; the
# escaped form renders as a plain "$" in HTML, PDF and DOCX and avoids that.
dollar_esc <- function(x) sub("$", "\\$", dollar(x, accuracy = 1), fixed = TRUE)
fmt_p <- function(p) {
  if (is.na(p)) return("NA")
  if (p < 0.001) "p < 0.001" else sprintf("p = %.3f", p)
}
# Inline reporting of a test statistic together with its p-value. wilcox_stat auto-labels the
# statistic V (paired signed-rank) or W (rank-sum); mcnemar_stat reports chi-squared with df.
# Math symbols ($V$, $W$, $\chi^2$) render in HTML, PDF and DOCX.
wilcox_stat  <- function(t) sprintf("$%s$ = %s, %s", names(t$statistic),
                                    comma(round(unname(t$statistic))), fmt_p(t$p.value))
mcnemar_stat <- function(t) sprintf("$\\chi^2$(%d) = %.1f, %s", unname(t$parameter),
                                    unname(t$statistic), fmt_p(t$p.value))

# Key paper-level values
Sys.setlocale("LC_TIME", "C")
month_year <- format(Sys.Date(), "%B %Y")
total_articles <- nrow(df_papers)
nih_articles   <- sum(df_papers$Funding_Status == "NIH Funded")
percent_funded <- nih_articles / total_articles
mean_cost          <- mean(df_papers$attributed_cost_adjusted, na.rm = TRUE)  # CPI-adjusted $
mean_cost_nominal  <- mean(df_papers$attributed_cost_raw,      na.rm = TRUE)  # nominal $
total_cost         <- sum(df_papers$attributed_cost_adjusted, na.rm = TRUE)   # CPI-adjusted $
total_cost_nominal <- sum(df_papers$attributed_cost_raw,      na.rm = TRUE)   # nominal $
# ===========================================================================
# Grant-level cost of NIH-funded retractions (beyond the per-paper attributed cost).
# Three complementary measures of direct cost are produced:
#   (1) total_cost              -> attributed cost: each grant's award divided by its number
#                                  of publications, summed over retracted papers [headline].
#   (2) total_cost_unique_grants-> full award of every DISTINCT grant behind a retracted
#                                  paper, counted ONCE (no per-paper attribution; no double-
#                                  counting of grants shared across papers).
#   (3) total_cost_personal     -> of (2), only grants awarded directly to the investigator
#                                  (research/career/fellowship + U01), excluding institutional
#                                  center/infrastructure grants (P30, M01, training, ...).
# Full rationale in the Supplementary methods.
# ===========================================================================

# --- (2) Unique-grant cost, no double counting -----------------------------
# Total nominal award of every DISTINCT NIH grant behind a retracted paper, each counted
# ONCE. A single grant can produce several retracted papers, so summing grant costs
# across paper rows would double-count shared grants.
#
# Inputs (both ";"-separated and aligned position-for-position on each NIH-funded paper row):
#   Funding_Info_cleaned       -> grant IDs (IC + serial), e.g. "CA203737;HL126711"
#   Individual_NIH_Grant_Costs -> the matching full awards, e.g. "5000000;2000000"
# We also carry the paper's CPI multiplier (Inflation_Multiplier, from its publication year)
# so the awards can be inflation-adjusted later.
.grant_costs <- df_papers %>%
  filter(Funding_Status == "NIH Funded") %>%
  # Split both fields and pair them by position into (grant, gcost), keeping the paper's mult.
  transmute(mult = Inflation_Multiplier,
            pair = map2(str_split(Funding_Info_cleaned, ";"),
                        str_split(Individual_NIH_Grant_Costs, ";"),
                        ~ tibble(grant = str_trim(.x), gcost = as.numeric(str_trim(.y))))) %>%
  unnest(pair) %>%                                   # -> one row per (paper, grant)
  filter(!is.na(grant), grant != "", !is.na(gcost)) %>%
  # Collapse to one row per DISTINCT grant (this is the de-duplication):
  #   gcost = first() -- a grant carries the SAME award in every paper that cites it (verified
  #                      internally consistent), so first() simply picks that single value.
  #   mult  = mean()  -- a grant may be cited by papers from several publication years, so we
  #                      use the mean of those papers' CPI multipliers as its representative
  #                      deflator (a grant has no single base year to deflate from).
  group_by(grant) %>% summarise(gcost = first(gcost), mult = mean(mult, na.rm = TRUE), .groups = "drop")
# Grants cited only by papers with no valid publication year get a non-finite mult; back-fill
# with the overall mean so they remain inflation-adjustable.
.mult_fallback <- mean(.grant_costs$mult[is.finite(.grant_costs$mult)])
.grant_costs   <- .grant_costs %>% mutate(mult = ifelse(is.finite(mult), mult, .mult_fallback))
n_unique_grants          <- nrow(.grant_costs)        # number of distinct grants
total_cost_unique_grants <- sum(.grant_costs$gcost)   # nominal $, each grant counted once

# --- (3) Personal vs institutional: recover each grant's NIH activity code --
# The activity code (R01, P30, U01, ...) is NOT in Funding_Info_cleaned but IS in the raw
# Funding_Info text, e.g. "NCI NIH HHS (R01 CA203737)". This regex captures the 3 parts of a
# grant number anywhere it appears: activity code | IC (2 letters) | serial. The optional
# space (\\s*) also matches concatenated forms such as "U19AI057229" or "5P50CA101942".
.code_pat <- "([A-Z][A-Z0-9]{2})\\s*([A-Z]{2})([0-9]{4,})"
# Map grant (IC+serial) -> activity code, parsed from the raw Funding_Info text.
.fi_map <- df_papers %>% filter(Funding_Status == "NIH Funded") %>%
  transmute(m = str_match_all(Funding_Info, .code_pat)) %>%            # every code per paper
  mutate(p = map(m, ~ tibble(grant = paste0(.x[, 3], .x[, 4]),         # grant = IC + serial
                             act   = .x[, 2]))) %>%                     # act   = activity code
  select(p) %>% unnest(p) %>%
  distinct(grant, act) %>% group_by(grant) %>% summarise(act_fi = first(act), .groups = "drop")
# Fallback map from the authors file, whose Grant_ID carries the full grant number
# (e.g. "K08HL086671"); recovers codes for grants not parseable from Funding_Info.
.au_map <- df_authors_raw %>%
  transmute(m = str_match(Grant_ID, .code_pat)) %>%
  transmute(grant = paste0(m[, 3], m[, 4]), act = m[, 2]) %>%
  filter(!is.na(act)) %>% distinct(grant, act) %>%
  group_by(grant) %>% summarise(act_au = first(act), .groups = "drop")
# Attach the code (prefer Funding_Info, else the authors file -> ~95% of dollars coded) and
# flag "personal" grants: research/career/fellowship (codes starting R/K/F) plus the NIH
# Director's high-risk personal awards (DP1/DP2/DP5). Excludes institutional career (K12/KL2),
# SBIR/STTR (R41-R44), institutional R-series mechanisms (R24 resource, R25 education, RL1
# linked), and U01 cooperative agreements (some are large multi-site consortia); P01 program
# projects are institutional here. Uncoded grants (NA) -> not personal.
.grants_typed <- .grant_costs %>%
  left_join(.fi_map, by = "grant") %>% left_join(.au_map, by = "grant") %>%
  mutate(activity = toupper(coalesce(act_fi, act_au)),
         personal = coalesce((str_detect(activity, "^[RKF]") &
                              !activity %in% c("K12", "KL2", "R24", "R25", "R41", "R42", "R43", "R44", "RL1")) |
                             activity %in% c("DP1", "DP2", "DP5"), FALSE))
total_cost_personal     <- sum(.grants_typed$gcost[.grants_typed$personal])                        # nominal $
total_cost_personal_adj <- sum((.grants_typed$gcost * .grants_typed$mult)[.grants_typed$personal]) # CPI-adjusted $
n_grants_personal       <- sum(.grants_typed$personal)
pct_personal_of_grants  <- total_cost_personal / total_cost_unique_grants * 100

# --- (4) Of the paper-linked personal grants, keep only those for which a retracted author is
# the PI. Author_Funding_Matches links each retracted author to the NIH grants they hold as PI
# (strict name match); we take those grant "cores" (IC + serial) and intersect them with the
# personal grants behind the retracted papers (so the retracted author is the grant holder,
# not merely acknowledged in the paper).
.pi_cores <- df_authors_raw %>%
  transmute(core = str_extract(Grant_ID, "[A-Z]{2}[0-9]{4,}$")) %>%
  filter(!is.na(core)) %>% distinct(core) %>% pull(core)
.personal_pi <- .grants_typed$personal & .grants_typed$grant %in% .pi_cores
total_cost_personal_pi     <- sum(.grants_typed$gcost[.personal_pi])                        # nominal $
total_cost_personal_pi_adj <- sum((.grants_typed$gcost * .grants_typed$mult)[.personal_pi]) # CPI-adjusted $
n_grants_personal_pi       <- sum(.personal_pi)                                             # distinct personal grants (PI-held)

# --- (5) NOT USED IN MANUSCRIPT: Career-wide investigator-held NIH funding of the retracted authors (as PI), regardless
# of whether a grant produced a retracted paper: the annual award of every distinct grant-year in
# which a retracted author is PI (de-duplicated across co-PIs), restricted to personal codes.
.career_gy <- df_authors_raw %>%
  filter(!is.na(Grant_ID), Grant_ID != "", !is.na(Funding_Amount)) %>%
  distinct(Grant_ID, Fiscal_Year, Funding_Amount) %>%
  left_join(annual_cpi, by = "Fiscal_Year") %>%
  mutate(adj = Funding_Amount * Inflation_Multiplier,
         act = str_extract(Grant_ID, "^[A-Z][A-Z0-9]{2}"),
         personal = coalesce((str_detect(act, "^[RKF]") &
                              !act %in% c("K12", "KL2", "R24", "R25", "R41", "R42", "R43", "R44", "RL1")) |
                             act %in% c("DP1", "DP2", "DP5"), FALSE))
total_career_personal     <- sum(.career_gy$Funding_Amount[.career_gy$personal], na.rm = TRUE)
total_career_personal_adj <- sum(.career_gy$adj[.career_gy$personal], na.rm = TRUE)

# Reference values for dynamic phrasing
nih_annual_budget   <- 47e9   # current NIH annual budget (USD)
stern2014_cost      <- 58e6   # USD reported in Stern et al. 2014
pct_of_budget       <- (total_cost / nih_annual_budget) * 100
fold_increase_stern <- total_cost / stern2014_cost

# Retraction lag (mean) + Mann-Whitney
valid_lag_data <- df_papers %>% filter(!is.na(retraction_lag) & retraction_lag >= 0)
nih_mean_lag_val <- round(mean(valid_lag_data$retraction_lag[valid_lag_data$Funding_Status == "NIH Funded"], na.rm = TRUE), 2)
oth_mean_lag_val <- round(mean(valid_lag_data$retraction_lag[valid_lag_data$Funding_Status == "Not NIH Funded"], na.rm = TRUE), 2)
nih_lag_test <- wilcox.test(
  valid_lag_data$retraction_lag[valid_lag_data$Funding_Status == "NIH Funded"],
  valid_lag_data$retraction_lag[valid_lag_data$Funding_Status == "Not NIH Funded"],
  exact = FALSE
)
nih_lag_p_txt <- wilcox_stat(nih_lag_test)

# % of NIH-funded papers in BLS or HSC
nih_domain_classified <- df_papers %>%
  filter(Funding_Status == "NIH Funded") %>%
  drop_na(Subject) %>%
  mutate(Paper_ID = row_number()) %>%
  separate_rows(Subject, sep = ";") %>%
  mutate(Subject = str_trim(Subject),
         Acronym = str_extract(Subject, "^\\([^)]+\\)")) %>%
  filter(Acronym %in% c("(BLS)", "(HSC)")) %>%
  distinct(Paper_ID)
nih_pct_bls_hsc <- n_distinct(nih_domain_classified$Paper_ID) / nih_articles * 100

# ===========================================================================
# CONSEQUENCES TO RETRACTED AUTHORS (Figure 1 and the "Consequences" section).
# Each author's NIH funding is tracked in a symmetric window around their FIRST
# retraction. Using diff_retracted = Fiscal_Year - First_Retraction_Year:
#   pre period  = years -3, -2, -1   |   post period = years +1, +2, +3.
# Year 0 (the retraction year itself) is excluded and belongs to neither period.
# All dollar amounts are inflation-adjusted (Funding_Adjusted). The cohort
# (df_authors_cohort) has already been (a) restricted to authors whose first
# retraction is early enough for all three post-years to be observable, and
# (b) name-collision-corrected at the grant level (see the grant_keep join above).
# ===========================================================================
total_cohort_authors <- n_distinct(df_authors_cohort$Original_Author_Name)

# Adjusted NIH funding per author per year offset (feeds the funded-author counts
# below and the within-author boxplots of Figure 1).
per_author_year_funding <- df_authors_cohort %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(Original_Author_Name, diff_retracted) %>%
  summarise(Annual_Funding = sum(Funding_Adjusted, na.rm = TRUE), .groups = "drop")

# Cohort-wide timeline: one row per year offset with the total adjusted funding,
# the number of distinct active grants, the count/share of authors holding any
# funding that year, and a funding index rebased to year -3 (= 100).
df_timeline <- df_authors_cohort %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(diff_retracted) %>%
  summarise(Total_Adjusted_Funding = sum(Funding_Adjusted, na.rm = TRUE),
            Active_Grants = n_distinct(Grant_ID[!is.na(Grant_ID)]),
            .groups = "drop") %>%
  left_join(
    per_author_year_funding %>%
      group_by(diff_retracted) %>%
      summarise(N_Funded = sum(Annual_Funding > 0), .groups = "drop"),
    by = "diff_retracted"
  ) %>%
  mutate(Cohort_Authors = total_cohort_authors,
         Average_Funding_Per_Author = Total_Adjusted_Funding / total_cohort_authors,
         Pct_With_Funding = N_Funded / total_cohort_authors * 100,
         Period_Label = ifelse(diff_retracted == 0, "Retraction Year (0)", "Normal Years"),
         Funding_Index = Average_Funding_Per_Author /
                         Average_Funding_Per_Author[diff_retracted == -3] * 100)

pct_funded_pre  <- mean(df_timeline$Pct_With_Funding[df_timeline$diff_retracted %in% c(-3, -2, -1)], na.rm = TRUE)
pct_funded_post <- mean(df_timeline$Pct_With_Funding[df_timeline$diff_retracted %in% c(1, 2, 3)], na.rm = TRUE)
pct_funded_change <- pct_funded_post - pct_funded_pre

total_annual_funding_pre  <- mean(df_timeline$Total_Adjusted_Funding[df_timeline$diff_retracted %in% c(-3, -2, -1)], na.rm = TRUE)
total_annual_funding_post <- mean(df_timeline$Total_Adjusted_Funding[df_timeline$diff_retracted %in% c(1, 2, 3)], na.rm = TRUE)

# Paired pre-vs-post table, one row per author. For each 3-year period we take the
# mean annual funding (period sum / 3) and the number of distinct active grants, then
# pivot to wide (Annual_Funding_Pre/Post, Total_Grants_Pre/Post). Authors with no
# funding rows in a period get 0 for that period (values_fill).
paired_data <- df_authors_cohort %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3 & diff_retracted != 0) %>%
  mutate(Period = ifelse(diff_retracted < 0, "Pre", "Post")) %>%
  group_by(Original_Author_Name, Period) %>%
  summarise(Annual_Funding = sum(Funding_Adjusted, na.rm = TRUE) / 3,
            Total_Grants  = n_distinct(Grant_ID[!is.na(Grant_ID)]),
            .groups = "drop") %>%
  pivot_wider(names_from = Period,
              values_from = c(Annual_Funding, Total_Grants),
              values_fill = list(Annual_Funding = 0, Total_Grants = 0))

# Add back the cohort authors who had NO funding rows anywhere in the window (absent
# from paired_data) as (0, 0), so every statistic below is over the FULL cohort.
paired_data_full <- df_authors_cohort %>% distinct(Original_Author_Name) %>%
  left_join(paired_data, by = "Original_Author_Name") %>%
  mutate(across(c(Annual_Funding_Pre, Annual_Funding_Post, Total_Grants_Pre, Total_Grants_Post),
                ~ ifelse(is.na(.x), 0, .x)))

# Cohort-wide central tendency of per-author annual funding, pre vs post. The median
# is highly sensitive to the large mass of $0 authors; the mean is reported alongside it.
median_pre  <- median(paired_data_full$Annual_Funding_Pre)
median_post <- median(paired_data_full$Annual_Funding_Post)
pct_change  <- if (median_pre > 0) (median_post - median_pre) / median_pre * 100 else NA_real_
# Cohort-wide MEAN per-author annual funding (robust to the zero mass, unlike the median).
mean_pre  <- mean(paired_data_full$Annual_Funding_Pre)
mean_post <- mean(paired_data_full$Annual_Funding_Post)

# Paired Wilcoxon signed-rank tests on the within-author pre-to-post change in annual
# funding and in active-grant count (non-parametric; the per-author distribution is
# strongly skewed and zero-inflated). wilcox_stat() formats "V = ..., p = ..." for inline text.
test_funding <- wilcox.test(paired_data_full$Annual_Funding_Pre, paired_data_full$Annual_Funding_Post,
                            paired = TRUE, exact = FALSE)
test_grants  <- wilcox.test(paired_data_full$Total_Grants_Pre,  paired_data_full$Total_Grants_Post,
                            paired = TRUE, exact = FALSE)
test_funding_p_txt <- wilcox_stat(test_funding)
test_grants_p_txt  <- wilcox_stat(test_grants)

# Change in the PROPORTION of authors holding any NIH funding (pre vs post) -> McNemar test.
test_funded <- mcnemar.test(table(
  factor(paired_data_full$Annual_Funding_Pre  > 0, levels = c(FALSE, TRUE)),
  factor(paired_data_full$Annual_Funding_Post > 0, levels = c(FALSE, TRUE))))
test_funded_p_txt <- mcnemar_stat(test_funded)

# lost_all_pct is the share of PREVIOUSLY FUNDED authors (Annual_Funding_Pre > 0) who hold no
# NIH funding in the 3 post-retraction years (not a share of the whole cohort).
n_funded_pre <- sum(paired_data_full$Annual_Funding_Pre > 0)
lost_all_n <- sum(paired_data_full$Annual_Funding_Pre > 0 & paired_data_full$Annual_Funding_Post == 0, na.rm = TRUE)
gained_n   <- sum(paired_data_full$Annual_Funding_Pre == 0 & paired_data_full$Annual_Funding_Post > 0, na.rm = TRUE)
lost_all_pct <- (lost_all_n / n_funded_pre) * 100
gained_pct   <- (gained_n   / total_cohort_authors) * 100

# Cohort-wide count of distinct active grants across the whole pre window vs the whole
# post window (a pooled count, not per author), and its percentage change.
total_grants_pre <- df_authors_cohort %>%
  filter(diff_retracted %in% c(-3, -2, -1)) %>%
  pull(Grant_ID) %>% na.omit() %>% n_distinct()
total_grants_post <- df_authors_cohort %>%
  filter(diff_retracted %in% c(1, 2, 3)) %>%
  pull(Grant_ID) %>% na.omit() %>% n_distinct()
pct_change_grants <- (total_grants_post - total_grants_pre) / total_grants_pre * 100

# ---------------------------------------------------------------------------
# Sensitivity subgroup: authors with more than one distinct retraction. Every
# object below is the "sens_"-prefixed counterpart of one computed above for the
# full cohort, using the same window, annualisation, zero-fill and tests.
# ---------------------------------------------------------------------------
sens_authors <- df_authors_cohort %>% filter(Num_Retractions > 1)
sens_cohort_authors <- n_distinct(sens_authors$Original_Author_Name)

per_author_year_funding_sens <- sens_authors %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(Original_Author_Name, diff_retracted) %>%
  summarise(Annual_Funding = sum(Funding_Adjusted, na.rm = TRUE), .groups = "drop")

sens_timeline <- sens_authors %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(diff_retracted) %>%
  summarise(Total_Adjusted_Funding = sum(Funding_Adjusted, na.rm = TRUE),
            Active_Grants = n_distinct(Grant_ID[!is.na(Grant_ID)]),
            .groups = "drop") %>%
  left_join(
    per_author_year_funding_sens %>%
      group_by(diff_retracted) %>%
      summarise(N_Funded = sum(Annual_Funding > 0), .groups = "drop"),
    by = "diff_retracted"
  ) %>% 
  mutate(Average_Funding_Per_Author = Total_Adjusted_Funding / sens_cohort_authors,
         Pct_With_Funding = N_Funded / sens_cohort_authors * 100,
         Period_Label = ifelse(diff_retracted == 0, "Retraction Year (0)", "Normal Years"),
         Funding_Index = Average_Funding_Per_Author /
                         Average_Funding_Per_Author[diff_retracted == -3] * 100)

sens_pct_funded_pre  <- mean(sens_timeline$Pct_With_Funding[sens_timeline$diff_retracted %in% c(-3, -2, -1)], na.rm = TRUE)
sens_pct_funded_post <- mean(sens_timeline$Pct_With_Funding[sens_timeline$diff_retracted %in% c(1, 2, 3)], na.rm = TRUE)
sens_pct_funded_change <- sens_pct_funded_post - sens_pct_funded_pre

sens_grants_pre <- sens_authors %>%
  filter(diff_retracted %in% c(-3, -2, -1)) %>%
  pull(Grant_ID) %>% na.omit() %>% n_distinct()
sens_grants_post <- sens_authors %>%
  filter(diff_retracted %in% c(1, 2, 3)) %>%
  pull(Grant_ID) %>% na.omit() %>% n_distinct()
sens_pct_change_grants <- (sens_grants_post - sens_grants_pre) / sens_grants_pre * 100

sens_paired_data <- sens_authors %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3 & diff_retracted != 0) %>%
  mutate(Period = ifelse(diff_retracted < 0, "Pre", "Post")) %>%
  group_by(Original_Author_Name, Period) %>%
  summarise(Annual_Funding = sum(Funding_Adjusted, na.rm = TRUE) / 3,
            Total_Grants  = n_distinct(Grant_ID[!is.na(Grant_ID)]),
            .groups = "drop") %>%
  pivot_wider(names_from = Period,
              values_from = c(Annual_Funding, Total_Grants),
              values_fill = list(Annual_Funding = 0, Total_Grants = 0))

sens_paired_data_full <- sens_authors %>% distinct(Original_Author_Name) %>%
  left_join(sens_paired_data, by = "Original_Author_Name") %>%
  mutate(across(c(Annual_Funding_Pre, Annual_Funding_Post, Total_Grants_Pre, Total_Grants_Post),
                ~ ifelse(is.na(.x), 0, .x)))

sens_median_pre  <- median(sens_paired_data_full$Annual_Funding_Pre)
sens_median_post <- median(sens_paired_data_full$Annual_Funding_Post)
sens_pct_change  <- if (sens_median_pre > 0) (sens_median_post - sens_median_pre) / sens_median_pre * 100 else NA_real_
sens_mean_pre  <- mean(sens_paired_data_full$Annual_Funding_Pre)
sens_mean_post <- mean(sens_paired_data_full$Annual_Funding_Post)

sens_test_funding <- wilcox.test(sens_paired_data_full$Annual_Funding_Pre, sens_paired_data_full$Annual_Funding_Post,
                                 paired = TRUE, exact = FALSE)
sens_test_grants  <- wilcox.test(sens_paired_data_full$Total_Grants_Pre,  sens_paired_data_full$Total_Grants_Post,
                                 paired = TRUE, exact = FALSE)
sens_test_funding_p_txt <- wilcox_stat(sens_test_funding)
sens_test_grants_p_txt  <- wilcox_stat(sens_test_grants)

# Change in the PROPORTION of authors holding any NIH funding (funded pre window vs post
# window) is a paired binary outcome -> McNemar test, to accompany the funded-fraction figures.
sens_test_funded <- mcnemar.test(table(
  factor(sens_paired_data_full$Annual_Funding_Pre  > 0, levels = c(FALSE, TRUE)),
  factor(sens_paired_data_full$Annual_Funding_Post > 0, levels = c(FALSE, TRUE))))
sens_test_funded_p_txt <- mcnemar_stat(sens_test_funded)

# Share of PREVIOUSLY FUNDED subgroup authors who lose all NIH funding post-retraction.
sens_n_funded_pre <- sum(sens_paired_data_full$Annual_Funding_Pre > 0)
sens_lost_all_n <- sum(sens_paired_data_full$Annual_Funding_Pre > 0 & sens_paired_data_full$Annual_Funding_Post == 0, na.rm = TRUE)
sens_gained_n   <- sum(sens_paired_data_full$Annual_Funding_Pre == 0 & sens_paired_data_full$Annual_Funding_Post > 0, na.rm = TRUE)
sens_lost_all_pct <- (sens_lost_all_n / sens_n_funded_pre) * 100
sens_gained_pct   <- (sens_gained_n   / sens_cohort_authors) * 100

add_margin <- function(v, pad = 0.10) c(0, max(v, na.rm = TRUE) * (1 + pad))
```

## ABSTRACT

Retractions attract substantial attention [@oransky2010] and have become more frequent over time [@sebo2025; @steen2016]. Retractions reflect the self-correcting nature of science but also wasted resources [@stern2014]. The Retraction Watch Database (RWD) includes over 60,000 records [@oransky2023; @sebo2025a]. We integrated RWD with NIH funding metadata (RePORTER system). As of `r month_year`, of the `r comma(total_articles)` U.S. affiliated retracted articles, `r comma(nih_articles)` (`r sprintf("%.1f%%", percent_funded * 100)`) were linked to at least one NIH grant. With a mean attributed cost per retracted NIH-funded article of `r dollar(mean_cost, accuracy = 1)` in `r cpi_year` dollars,  the attributed total cost of NIH-funded retracted research is `r dollar(total_cost, scale = 1e-6, accuracy = 1, suffix = " million")` in `r cpi_year` dollars. Grants associated with retracted papers for which the first or last author of the paper was the principal investigator were awarded `r dollar(total_cost_personal_pi_adj, scale = 1e-9, accuracy = 0.01, suffix = " billion")` in `r cpi_year` dollars. NIH-funded articles take longer to be retracted (mean = `r nih_mean_lag_val` years) than other US-based articles, which may entail greater downstream implications. NIH funding of retracted authors decreased over the 3 years following retraction, particularly among authors with multiple retractions. We have developed a continuously updated dynamic dashboard (<https://sandovallentisco.shinyapps.io/nih-retractions/>) for the cost of retractions reflecting NIH-funded work.


Retracted publications are a visible manifestation of how science self-corrects, but they may also signal serious weaknesses in research practices and peer review [@oransky2010]. Retracted work, especially when it is due to serious fraudulent practices or major flaws and not just publisher or journal technical errors, may undermine trust in science, distort scientific evidence, and represent a waste of scarce research resources. Although retractions remain rare relative to overall publication volume, their frequency has been climbing for two decades [@sebo2025; @steen2016]. One would wish to estimate the cost of the research work that has been retracted. Previous estimates of the cost of retracted research for the U.S. National Institutes of Health (NIH), the world's largest public funder of biomedical research, examined 149 retracted NIH-funded papers between 1992 and 2012 and reported a direct cost of about \$58 million [@stern2014], but this figure is now outdated. Far more comprehensive accounting is possible today using the Retraction Watch Database (RWD), a curated, regularly updated catalogue of more than 60,000 retraction records across all disciplines [@oransky2023; @sebo2025a], and NIH funding metadata, accessible through the agency's RePORTER system. We integrated the two systems via an automated weekly pipeline and developed a dynamic dashboard that updates continuously as new retractions are cataloged (<https://sandovallentisco.shinyapps.io/nih-retractions/>). We present here the data for the cost of retracted NIH-funded research as of `r month_year` and discuss some implications of these findings.

## Direct cost to the NIH

We retrieved the latest version of the Retraction Watch Database (RWD) from the Crossref repository (<https://gitlab.com/crossref/retraction-watch-data>) and selected entries marked as retractions, excluding expressions of concern or corrections. We also excluded entries classified as ‘Retract and Replace’, and retractions that did not represent a high likelihood of serious error and/or misconduct by the authors (e.g., administrative or publisher errors). For those eligible retracted articles for which RWD identified an institutional affiliation within the U.S., we queried PubMed using the National Center for Biotechnology Information (NCBI) Entrez API to get supplementary funding metadata (grant IDs and number of publications linked to each grant). Finally, to retrieve the funding amount associated with each grant, we used the NIH Research Portfolio Online Reporting Tools Expenditure and Results (RePORTER) tool. Databases from all available fiscal years were downloaded using the ExPORTER tool (see more details about the method in Supplementary Materials). Data and scripts are available at <https://github.com/sandovallentisco/nih-retractions-cost>.

As of `r month_year`, the RWD listed `r comma(total_articles)` retracted articles with at least one U.S. affiliated author. Of these, `r comma(nih_articles)` (`r sprintf("%.1f%%", percent_funded * 100)`) were linked to at least one NIH grant. To calculate the attributed cost per retracted publication, we divided the total grant award by the total number of publications produced under that grant (i.e., attributing a proportional cost to the specific retracted article rather than the entire grant value). The mean attributed cost per retracted NIH-funded article was `r dollar(mean_cost_nominal, accuracy = 1)` in nominal dollars, `r dollar(mean_cost, accuracy = 1)` in `r cpi_year` dollars. Summed across all papers, the attributed total direct cost is `r dollar(total_cost_nominal, scale = 1e-6, accuracy = 1, suffix = " million")` in nominal dollars, `r dollar(total_cost, scale = 1e-6, accuracy = 1, suffix = " million")` in `r cpi_year` dollars, about 1% of a single year of NIH budget, but a `r sprintf("%.1f", fold_increase_stern)`-fold increase over the previously published benchmark that covered retractions during 1992-2012 [@stern2014]. The retracted publications were associated with a total of `r comma(n_unique_grants)` distinct NIH grants, i.e., a total of `r dollar_short(total_cost_unique_grants)` in nominal dollars. However, the vast majority of this sum is institutional center and infrastructure funding (e.g., P30 center core grants) rather than money awarded to individual investigators. Considering only investigator-held research, career and fellowship awards (R/K/F-series; see Supplementary materials for the exact activity-code rule) for which the retracted author was the principal investigator, a total amount of `r dollar_short(total_cost_personal_pi)` in nominal dollars or `r dollar_short(total_cost_personal_pi_adj)` in `r cpi_year` dollars from `r comma(n_grants_personal_pi)` grants was associated with retracted papers.

Retracted articles only reflect the work the literature has managed to identify and withdraw, but not the much larger body of flawed or fraudulent work that has yet to be exposed or has been exposed but not yet retracted [@naddaf2026; @scancar2026; @topaz2026]. Therefore, the `r dollar(total_cost, scale = 1e-6, accuracy = 1, suffix = " million")` is clearly an underestimate of the wasted resources. The `r dollar_short(total_cost_personal_pi_adj)` can be either an overestimate or an underestimate of the wasted resources. Much of these researchers’ other work may have been without serious flaws; conversely, other work by other researchers who have no retraction yet may have been seriously flawed or even fraudulent.

## Consequences to retracted authors

We tracked the funding trajectories of `r comma(total_cohort_authors)` first/last authors for the 3 pre- and post-retraction years. We selected cases in which the first retraction occurred by `r retraction_cutoff_year`, so that a full 3-year post-retraction window of funding data was observable for every author in the cohort given the most recent RePORTER fiscal year (`r latest_fiscal_year`). Because common author names can match several different NIH investigators, for the `r comma(n_collision_authors)` authors whose name maps to more than one investigator we kept only the grants of the one whose institution matches their retracted paper (see Supplementary materials for details). We observed some potential decrease in NIH funding for retracted authors. The per-author cohort-wide median annual NIH funding fell from `r dollar_esc(median_pre)` in the 3 years pre-retraction to `r dollar_esc(median_post)` in the 3 years post-retraction, a `r sprintf("%+.1f%%", pct_change)` change (paired Wilcoxon `r test_funding_p_txt`). This decrease in the median is due to a higher proportion of authors receiving no funding after retraction; the average annual fraction of the cohort holding an active NIH grant fell from `r sprintf("%.1f%%", pct_funded_pre)` to `r sprintf("%.1f%%", pct_funded_post)` (McNemar's test `r test_funded_p_txt`). However, the mean remained very similar (`r dollar_esc(mean_pre)` pre vs `r dollar_esc(mean_post)` post; the full per-author funding distribution before and after retraction is shown in the Supplementary Results). The number of unique active grants dropped by `r sprintf("%.1f%%", abs(pct_change_grants))` (paired Wilcoxon `r test_grants_p_txt`). `r sprintf("%.1f%%", lost_all_pct)` of previously funded authors lost all NIH funding in the 3 years after their first retraction.

Changes were more pronounced for authors with more than one retracted paper (n = `r comma(sens_cohort_authors)`). Their median annual funding fell from `r dollar_esc(sens_median_pre)` to `r dollar_esc(sens_median_post)` (`r sprintf("%+.1f%%", sens_pct_change)`; paired Wilcoxon `r sens_test_funding_p_txt`), and the mean annual fraction of the cohort holding an active NIH grant fell from `r sprintf("%.1f%%", sens_pct_funded_pre)` to `r sprintf("%.1f%%", sens_pct_funded_post)` (McNemar's test `r sens_test_funded_p_txt`). In contrast to the full cohort, here the mean also decreased substantially (`r dollar_esc(sens_mean_pre)` pre vs `r dollar_esc(sens_mean_post)` post) (See Fig. 1). The number of unique active grants dropped by `r sprintf("%.1f%%", abs(sens_pct_change_grants))` (paired Wilcoxon `r sens_test_grants_p_txt`), and `r sprintf("%.1f%%", sens_lost_all_pct)` of previously funded authors in this subgroup lost all NIH funding in the 3 years after their first retraction.

```{r fig1-policy, fig.width=12, fig.height=13, dpi=300}
# FIGURE 1. A 2x2 grid of decomposed boxplots showing, for every year offset -3..+3,
# the within-author CHANGE relative to baseline year -3. Top row = full cohort, bottom
# row = the >1-retraction subgroup; left column = NIH funding, right column = active-grant
# count. Built below in stages: (1) a small theme; (2) make_diff_box_data() computes the
# per-author differences vs year -3; (3) box_stats()/mean_line()/whisker_cap() derive the
# boxplot geometry; (4) make_box_panel() draws one panel; (5) the four panels, two row-title
# strips and a manual legend are assembled with patchwork.
#
# Font sizes matched to Figure 2 (dashboard_theme): base 18, axis titles 20, axis text 18.
small_theme <- theme_classic(base_size = 18) +
  theme(
    legend.position   = "none",
    axis.title        = element_text(size = 20, face = "bold"),
    axis.title.y      = element_text(size = 17, face = "bold"),
    axis.title.x      = element_text(size = 17, face = "bold"),
    axis.text         = element_text(size = 18),
    plot.background   = element_rect(fill = "transparent", color = NA),
    panel.background  = element_rect(fill = "transparent", color = NA),
    panel.grid.major.y = element_line(color = "gray90", linewidth = 0.5)
  )

tl_fig2      <- df_timeline
tl_fig2_sens <- sens_timeline
timeline_colors_fig2 <- c("Retraction Year (0)" = "#cccccc", "Normal Years" = "#005b96")

y_limits_grants       <- add_margin(tl_fig2$Active_Grants)
y_limits_grants_sens  <- add_margin(tl_fig2_sens$Active_Grants)

# Panels A-D: within-author change vs. the baseline year -3. For each author and year offset,
# the difference between their value that year (funding dollars for A/C, active-grant count for
# B/D) and their value at year -3. A within-author subtraction (not a ratio), so it is defined
# for the FULL cohort with no restriction, and year -3 is identically 0. (See the preprint's
# Supplementary methods for the full rationale.)
make_diff_box_data <- function(per_author_value, cohort_names, value_col = "Annual_Funding") {
  grid <- tidyr::crossing(Original_Author_Name = cohort_names, diff_retracted = -3:3) %>%
    left_join(per_author_value, by = c("Original_Author_Name", "diff_retracted")) %>%
    mutate(.val = ifelse(is.na(.data[[value_col]]), 0, .data[[value_col]]))
  baseline_by_author <- grid %>% filter(diff_retracted == -3) %>%
    transmute(Original_Author_Name, Baseline = .val)
  grid %>% left_join(baseline_by_author, by = "Original_Author_Name") %>%
    mutate(Funding_Diff = .val - Baseline,
           Period_Label = ifelse(diff_retracted == 0, "Retraction Year (0)", "Normal Years"))
}
# Per-author per-year count of distinct active NIH grants (for Panels B/D).
per_author_year_grants <- df_authors_cohort %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(Original_Author_Name, diff_retracted) %>%
  summarise(N_Grants = n_distinct(Grant_ID[!is.na(Grant_ID)]), .groups = "drop")
per_author_year_grants_sens <- sens_authors %>%
  filter(diff_retracted >= -3 & diff_retracted <= 3) %>%
  group_by(Original_Author_Name, diff_retracted) %>%
  summarise(N_Grants = n_distinct(Grant_ID[!is.na(Grant_ID)]), .groups = "drop")

box_fig2        <- make_diff_box_data(per_author_year_funding,      unique(df_authors_cohort$Original_Author_Name))
box_fig2_sens   <- make_diff_box_data(per_author_year_funding_sens, unique(sens_authors$Original_Author_Name))
box_grants      <- make_diff_box_data(per_author_year_grants,       unique(df_authors_cohort$Original_Author_Name), "N_Grants")
box_grants_sens <- make_diff_box_data(per_author_year_grants_sens,  unique(sens_authors$Original_Author_Name),       "N_Grants")

# Boxplot statistics on the differences: box = IQR, whiskers = 10th-90th percentiles, centre = median.
box_stats <- function(box_data) box_data %>%
  group_by(diff_retracted, Period_Label) %>%
  summarise(p10 = quantile(Funding_Diff, 0.10), p25 = quantile(Funding_Diff, 0.25),
            p50 = median(Funding_Diff), p75 = quantile(Funding_Diff, 0.75),
            p90 = quantile(Funding_Diff, 0.90), .groups = "drop")
stat_fig2      <- box_stats(box_fig2)
stat_fig2_sens <- box_stats(box_fig2_sens)

# Mean change overlaid as a trajectory (box centre is the median, also in dollars).
mean_line <- function(box_data) box_data %>%
  group_by(diff_retracted) %>% summarise(Funding_Diff = mean(Funding_Diff), .groups = "drop")
mean_fig2      <- mean_line(box_fig2)
mean_fig2_sens <- mean_line(box_fig2_sens)

# Outliers beyond the 10th-90th percentile whiskers.
outlier_points <- function(box_data) box_data %>%
  group_by(diff_retracted) %>%
  mutate(.p10 = quantile(Funding_Diff, 0.10), .p90 = quantile(Funding_Diff, 0.90)) %>%
  ungroup() %>% filter(Funding_Diff < .p10 | Funding_Diff > .p90)
out_fig2      <- outlier_points(box_fig2)
out_fig2_sens <- outlier_points(box_fig2_sens)

# Grant panels (B/D): two-sided display cap at the 2.5th-97.5th pct of the differences.
diff_cap <- function(box_data)
  c(quantile(box_data$Funding_Diff, 0.025, na.rm = TRUE),
    quantile(box_data$Funding_Diff, 0.975, na.rm = TRUE))
# Tight axis fit to the visible elements (10th-90th-pct whiskers + mean), with only a small 5%
# margin so the whiskers nearly fill the panel and the within-author changes read more clearly
# (outliers are not drawn). Used for all four panels.
whisker_cap <- function(stat_data, mean_data) {
  lo <- min(c(stat_data$p10, mean_data$Funding_Diff), na.rm = TRUE)
  hi <- max(c(stat_data$p90, mean_data$Funding_Diff), na.rm = TRUE)
  pad <- 0.05 * (hi - lo)
  c(lo - pad, hi + pad)
}
y_cap_funding      <- whisker_cap(stat_fig2,      mean_fig2)
y_cap_funding_sens <- whisker_cap(stat_fig2_sens, mean_fig2_sens)

# Same statistics for the active-grant-count differences (Panels B and D).
stat_grants <- box_stats(box_grants);      stat_grants_sens <- box_stats(box_grants_sens)
mean_grants <- mean_line(box_grants);      mean_grants_sens <- mean_line(box_grants_sens)
out_grants  <- outlier_points(box_grants); out_grants_sens  <- outlier_points(box_grants_sens)
y_cap_grants <- whisker_cap(stat_grants, mean_grants); y_cap_grants_sens <- whisker_cap(stat_grants_sens, mean_grants_sens)

# (Unused) line-plot factory, retained for reference.
make_panel <- function(data, y, y_label, y_limits, dollar_y = TRUE,
                       title = NULL, hline_at = NULL) {
  p <- ggplot(data, aes(x = diff_retracted, y = .data[[y]])) +
    geom_vline(xintercept = 0, color = "#d9534f", linetype = "dashed", linewidth = 0.8)
  if (!is.null(hline_at))
    p <- p + geom_hline(yintercept = hline_at, color = "gray60", linetype = "dotted", linewidth = 0.6)
  p <- p +
    geom_line(color = "#005b96", linewidth = 0.5, group = 1) +
    geom_point(aes(color = Period_Label), size = 2.5) +
    scale_color_manual(values = timeline_colors_fig2) +
    scale_x_continuous(breaks = -3:3) +
    scale_y_continuous(labels = if (dollar_y) label_dollar() else comma,
                       limits = y_limits, expand = expansion(mult = c(0, 0))) +
    labs(x = "Years from First Retraction", y = y_label, title = title) +
    small_theme +
    theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())
  if (!is.null(title)) p <- p + theme(plot.title = element_text(size = 14, face = "bold", hjust = 0))
  p
}

# Boxplot factory for all four panels: within-author change vs. year -3. Drawn as separate
# layers so the IQR box has NO outline -- a thick errorbar for the 10th-90th-pct whiskers (with
# T-caps), a filled tile for the IQR (no border), and a segment for the median. Boxes are
# coloured by period (green = pre-retraction -3..0, orange = post 1..3); the mean change is a
# continuous black line with black points. No dotted zero line; no y-axis line/ticks.
make_box_panel <- function(stat_data, mean_data, outlier_data, y_cap, title = NULL,
                           y_label = "Change in annual NIH funding", dollar_y = TRUE) {
  period_pal <- c("Pre" = "#00716f", "Post" = "#d16b2a")
  bw <- 0.6
  stat_data <- stat_data %>% mutate(Period2 = ifelse(diff_retracted <= 0, "Pre", "Post"))
  mean_data <- mean_data %>% mutate(Period2 = ifelse(diff_retracted <= 0, "Pre", "Post"))
  p <- ggplot(stat_data, aes(x = diff_retracted)) +
    geom_vline(xintercept = 0, color = "#d9534f", linetype = "dashed", linewidth = 0.8) +
    geom_errorbar(aes(ymin = p10, ymax = p90), width = 0.3, linewidth = 0.4, color = "gray30") +
    geom_tile(aes(y = (p25 + p75) / 2, height = p75 - p25, fill = Period2),
              width = bw, color = NA) +
    geom_segment(aes(x = diff_retracted - bw / 2, xend = diff_retracted + bw / 2,
                     y = p50, yend = p50), color = "gray30", linewidth = 0.5) +
    geom_line(data = mean_data, aes(x = diff_retracted, y = Funding_Diff),
              color = "black", linewidth = 0.8, group = 1) +
    geom_point(data = mean_data, aes(x = diff_retracted, y = Funding_Diff),
               color = "black", size = 2.2) +
    scale_fill_manual(values = period_pal) +
    scale_x_continuous(breaks = -3:3) +
    scale_y_continuous(labels = if (dollar_y) label_dollar() else label_number(accuracy = 1)) +
    coord_cartesian(ylim = y_cap) +
    labs(x = "Years from First Retraction", y = y_label, title = title) +
    small_theme +
    theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())
  if (!is.null(title)) p <- p + theme(plot.title = element_text(size = 20, face = "bold", hjust = 0))
  p
}

p_A <- make_box_panel(stat_fig2,      mean_fig2,      out_fig2,      y_cap_funding)
p_B <- make_box_panel(stat_grants,      mean_grants,      out_grants,      y_cap_grants,
                      y_label = "Change in active NIH grants", dollar_y = FALSE)
p_C <- make_box_panel(stat_fig2_sens, mean_fig2_sens, out_fig2_sens, y_cap_funding_sens)
p_D <- make_box_panel(stat_grants_sens, mean_grants_sens, out_grants_sens, y_cap_grants_sens,
                      y_label = "Change in active NIH grants", dollar_y = FALSE)

# Row titles drawn as full-width text strips, placed ABOVE each panel row (above the A/B and
# C/D tags) so each title spans its two panels.
title_strip <- function(txt) ggplot() +
  annotate("text", x = 0.5, y = 0.5, label = txt, fontface = "bold", size = 7) +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1), clip = "off") +
  theme_void()
title1_plot <- title_strip(paste0("All retracted authors (N = ", comma(total_cohort_authors), ")"))
title2_plot <- title_strip(paste0("Authors with >1 retracted paper (N = ", comma(sens_cohort_authors), ")"))

# Manual legend (bottom strip): period colours, the mean overlay, and the boxplot anatomy.
legend_plot <- ggplot() +
  coord_cartesian(xlim = c(0, 100), ylim = c(0, 100), clip = "off") +
  # Row 1: period colours (well separated to avoid overlap)
  annotate("rect", xmin = 2, xmax = 5, ymin = 66, ymax = 80, fill = "#00716f", color = NA) +
  annotate("text", x = 6.5, y = 73, hjust = 0, size = 5.2, label = "Pre-retraction (years -3 to 0)") +
  annotate("rect", xmin = 52, xmax = 55, ymin = 66, ymax = 80, fill = "#d16b2a", color = NA) +
  annotate("text", x = 56.5, y = 73, hjust = 0, size = 5.2, label = "Post-retraction (years 1 to 3)") +
  # Row 2: mean marker + boxplot anatomy
  annotate("segment", x = 2, xend = 8, y = 26, yend = 26, color = "black", linewidth = 0.9) +
  annotate("point", x = 5, y = 26, color = "black", size = 3) +
  annotate("text", x = 9.5, y = 26, hjust = 0, size = 5.2, label = "Mean change") +
  annotate("segment", x = 33, xend = 33, y = 8, yend = 44, color = "gray30", linewidth = 0.7) +
  annotate("segment", x = 31.8, xend = 34.2, y = 8,  yend = 8,  color = "gray30", linewidth = 0.7) +
  annotate("segment", x = 31.8, xend = 34.2, y = 44, yend = 44, color = "gray30", linewidth = 0.7) +
  annotate("rect", xmin = 31.4, xmax = 34.6, ymin = 18, ymax = 34, fill = "gray80", color = NA) +
  annotate("segment", x = 31.4, xend = 34.6, y = 26, yend = 26, color = "gray30", linewidth = 0.6) +
  annotate("text", x = 36.5, y = 26, hjust = 0, size = 5.2,
           label = "Box = IQR;  centre line = median;  whiskers = 10th-90th pct") +
  theme_void()

# Title strip over each panel row, the 2x2 boxplot grid, then the legend. Area letters run in
# alphabetical = plot order; only the four data panels carry a tag (A-D).
design <- "AAAA\nBBCC\nDDDD\nEEFF\nGGGG"
(title1_plot + p_A + p_B + title2_plot + p_C + p_D + legend_plot) +
  plot_layout(design = design, heights = c(0.14, 1.2, 0.14, 1.2, 0.28)) +
  plot_annotation(tag_levels = list(c("", "A", "B", "", "C", "D", "")))
```

**Figure 1. Change in funding for retracted authors compared to year -3** Top row (A, B): the full cohort of `r comma(total_cohort_authors)` retracted first/last authors; bottom row (C, D): the `r comma(sens_cohort_authors)` authors with more than one retracted paper. (A, C) inflation-adjusted NIH funding; (B, D) count of distinct active NIH grants. Every panel shows the within-author change relative to the baseline year −3 as a boxplot.

## Time to retract NIH-funded articles

Time-to-retraction (see Fig. 2) is markedly longer for NIH-funded papers than for U.S. retractions without NIH funding (`r nih_mean_lag_val` versus `r oth_mean_lag_val` years on average; Mann-Whitney rank-sum, `r nih_lag_p_txt`). This gap is largely explained by the different mix of retraction reasons: NIH-funded papers are retracted more often for image-integrity and formal misconduct issues (e.g., data fabrication), which take longer to surface and resolve, whereas non-NIH papers are retracted more often for plagiarism, paper mills, or compromised peer review which may be faster and more definitive to document and retract (see more details in Supplementary Table 1 in Supplementary Materials). The practical implication is that NIH-funded results spend longer in the literature before being retracted, with more time to seed downstream studies, clinical practice, and subsequent funding decisions.

```{r}
#| fig-width: 12
#| fig-height: 10
#| warning: false

# FIGURE 2. Three stacked-bar panels coloured by funding status: (A) retracted articles
# by publication year, (B) by retraction year, and (C) the time-to-retraction (lag)
# distribution with the group mean lags overlaid as dashed lines. Panels A/B drop pre-1980
# papers and Panel C caps the x-axis at LAG_AXIS_MAX years for readability (counts reported
# in the caption). Assembled with patchwork; the group means come from the setup chunk.
n_excluded_pub <- sum(df_papers$Publication_Year < 1980, na.rm = TRUE)
n_excluded_ret <- sum(df_papers$Retraction_Year < 1980, na.rm = TRUE)

p_pub_year <- df_papers %>% drop_na(Publication_Year) %>% filter(Publication_Year >= 1980) %>%
  rename(`Publication Year` = Publication_Year, `Funding Status` = Funding_Status) %>%
  count(`Publication Year`, `Funding Status`, name = "No. of articles") %>%
  ggplot(aes(x = `Publication Year`, y = `No. of articles`, fill = `Funding Status`)) +
  geom_col(width = 1, color = "white", position = "stack") +
  scale_fill_manual(values = corporate_colors) +
  labs(x = "Publication Year", y = "No. of articles") + dashboard_theme +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())

p_ret_year <- df_papers %>% drop_na(Retraction_Year) %>% filter(Retraction_Year >= 1980) %>%
  rename(`Retraction Year` = Retraction_Year, `Funding Status` = Funding_Status) %>%
  count(`Retraction Year`, `Funding Status`, name = "No. of articles") %>%
  ggplot(aes(x = `Retraction Year`, y = `No. of articles`, fill = `Funding Status`)) +
  geom_col(width = 1, color = "white", position = "stack") +
  scale_fill_manual(values = corporate_colors) +
  labs(x = "Retraction Year", y = "No. of articles") + dashboard_theme +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())

LAG_AXIS_MAX <- 25

n_total_lag   <- nrow(df_papers)
n_missing_lag <- sum(is.na(df_papers$retraction_lag))
n_negative    <- sum(df_papers$retraction_lag < 0, na.rm = TRUE)
valid_lag     <- df_papers %>% filter(!is.na(retraction_lag) & retraction_lag >= 0)
n_outliers    <- sum(valid_lag$retraction_lag > LAG_AXIS_MAX, na.rm = TRUE)
# Tallest visible bar, used to position the means key box in Panel C
ymax_lag <- valid_lag %>% filter(retraction_lag <= LAG_AXIS_MAX) %>%
  count(retraction_lag) %>% summarise(m = max(n, na.rm = TRUE)) %>% pull(m)

p_lag <- valid_lag %>%
  rename(`Retraction Lag` = retraction_lag, `Funding Status` = Funding_Status) %>%
  count(`Retraction Lag`, `Funding Status`, name = "No. of articles") %>%
  ggplot(aes(x = `Retraction Lag`, y = `No. of articles`, fill = `Funding Status`)) +
  geom_col(width = 1, color = "white", position = "stack") +
  # Mean retraction lag for each group, shown on Panel C (Panel D removed)
  geom_vline(xintercept = oth_mean_lag_val, color = "#00716f", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = nih_mean_lag_val, color = "#d16b2a", linetype = "dashed", linewidth = 1) +
  # Means key: framed box with a dashed colour sample per group and black text labels
  annotate("rect", xmin = 12, xmax = LAG_AXIS_MAX, ymin = 0.73 * ymax_lag, ymax = 1.0 * ymax_lag,
           fill = "white", color = "white", linewidth = 0.5) +
  annotate("segment", x = 12.7, xend = 15.2, y = 0.92 * ymax_lag, yend = 0.92 * ymax_lag,
           color = "#d16b2a", linetype = "dashed", linewidth = 1.1) +
  annotate("segment", x = 12.7, xend = 15.2, y = 0.80 * ymax_lag, yend = 0.80 * ymax_lag,
           color = "#00716f", linetype = "dashed", linewidth = 1.1) +
  annotate("text", x = 15.7, y = 0.92 * ymax_lag, hjust = 0, color = "black", size = 4.7,
           label = sprintf("NIH mean = %.1f yr", nih_mean_lag_val)) +
  annotate("text", x = 15.7, y = 0.80 * ymax_lag, hjust = 0, color = "black", size = 4.7,
           label = sprintf("Non-NIH mean = %.1f yr", oth_mean_lag_val)) +
  scale_fill_manual(values = corporate_colors) +
  coord_cartesian(xlim = c(0, LAG_AXIS_MAX)) +
  labs(x = "Years between Publication and Retraction", y = "No. of articles") +
  dashboard_theme +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())

# Layout: A and B on the top row; C (the lag distribution, now annotated with the group
# means) centred on the bottom row. "#" are empty cells that centre Panel C.
design <- "AABB\n#CC#"
combined_plot <- p_pub_year + p_ret_year + p_lag +
  plot_layout(design = design, guides = "collect") &
  theme(legend.position = "top")

combined_plot <- combined_plot + plot_annotation(tag_levels = 'A')

print(combined_plot)
```

**Figure 2. Publication and Retraction Trends.** (A) Distribution of publication years. (B) Distribution of retraction years. (C) Lag between publication and retraction; dashed lines mark the mean lag for NIH-funded (orange) and non-NIH (green) papers. Note: For readability, Panels A and B exclude `r n_excluded_pub` papers published and `r n_excluded_ret` papers retracted before 1980. Likewise, the X axis of Panel C is capped at `r LAG_AXIS_MAX` years; `r format(n_outliers, big.mark = ",")` papers with a lag greater than `r LAG_AXIS_MAX` years are excluded from the plot but included in the means.

## Additional considerations and implications

According to our estimates, the direct financial costs of NIH-funded retracted only represent a very small portion of the NIH budget. However, far more papers are seriously flawed, misleading, or even fraudulent and never get retracted. Therefore, the wasted resources may be much larger. We also found that investigator-held grants for which the retracted author was the principal investigator were awarded `r dollar_short(total_cost_personal_pi_adj)` in `r cpi_year` dollars. Therefore, the footprint of flawed and fraudulent work may relate to a much larger space of research efforts. Furthermore, retraction mechanisms remain slow and inefficient as of today, and they only reflect the visible part of a problem that might be much bigger [@naddaf2026; @scancar2026; @topaz2026]. Besides, direct cost is also an underestimate of damage since downstream research that built on flawed conclusions or clinical decisions informed by them also entail cost. Evidence synthesis may also be affected [@bakker2024].

Enhancing detection of papers that warrant retraction is not a simple process. Mass-detection tools (image forensics, "tortured-phrase" scanners, statistical anomaly detectors, or emerging AI-content classifiers) may be funded and developed as research infrastructure and applied as pre- and post-publication surveillance. Efficiency and yield of different tools require careful study. Accurate and robust methods to detect fraudulent papers might prevent their publication if applied at the peer review stage or, at least, should enable shorter retraction times, which is especially relevant considering the longer time-to-retraction for NIH-funded studies. Moreover, it should be acknowledged that many retractions reflect honest errors and their authors who detect them and withdraw their flawed work from the published corpus should be congratulated for their diligence and honesty. Creating a culture where honest correction or error is rewarded rather than leading to reputational damage is warranted [@fanelli2018; @hosseini2018; @fanelli2016]. Currently, most retractions probably reflect poor research practices and misconduct [@hwang2023], but the balance may change if more honest error retractions are incentivized.  

Another important point is that retracted authors experienced a short-term decrease in NIH funding, with the most pronounced declines among authors with more than one retracted article, although a big proportion of previously funded authors still retained at least some funding. One may debate whether and how the NIH might consider enforcing faster and stricter consequences for retracted authors, especially when the reasons for retractions are not due to honest errors, but to serious fraudulent practices such as data fabrication. "Super-retractor" authors with a large number of retractions may account for a sizeable share of total retractions [@lyu2026] and may be a particular group warranting more aggressive measures. Some of these authors may be highly cited and may have a major negative impact in their fields [@ioannidis2025].

Lastly, considering the fast pace with which the published scientific literature is evolving, static reports quickly become obsolete. The dashboard accompanying this analysis (<https://sandovallentisco.shinyapps.io/nih-retractions/>) is automatically recomputed weekly from openly available data, so the estimates in this article will continue to refresh after publication. We encourage meta-scientists and policy researchers to also include living, automated analyses of research practices and research outcomes when possible.

## Supplementary materials

Detailed methods and additional analyses can be found at: <https://github.com/sandovallentisco/nih-retractions-cost/blob/main/supplementary_materials.pdf>

## Disclosures and data availability statement

This study was preregistered on March 23, 2026, on OSF (<https://osf.io/guq59/>). Deviations from preregistration can be found at <https://osf.io/kd8bf/>. All data and scripts can be found at <https://github.com/sandovallentisco/nih-retractions-cost>.

## References

::: {#refs}
:::
